Astro 507; Problem Set 2

### 2022-01-28 pset2/solutions.tex
minor changes
reset q1
use capital phi
finish off question 1

\question{\textbf{1. Heat Capacities}}
\answer{
    The definition of heat capacity gives that
    \begin{equation}
        C \equiv \dv{Q}{T}
    \end{equation}
    $C_P$ and $C_V$ are $C$ but with the pressure, $P$, and volume, $V$, each held constant respectively. Applying the first law of thermodynamics, this gives
    \begin{equation}
        C_V = \dv{U}{T},\qquad C_P = \dv{U}{T} + P \dv{V}{T}
    \end{equation}
    Therefore the difference of the two is found as follows (applying the ideal gas law).
    \begin{align}
        C_P - C_V &= P \dv{V}{T} \\
                  &= P \dv{T} \qty(\frac{N k_B T}{P}) \\
                  &= P \qty(\frac{N k_B}{P}) \\
        \Aboxed{ C_P - C_V &= N k_B }
    \end{align}
    Now we can additionally find the ratio of the two
    \begin{align}
        \frac{C_P}{C_V} &= \frac{\dv{U}{T} + P \dv{V}{T}}{\dv{U}{T}} \\
                        &= 1 + P \dv{V}{T} \dv{T}{U} \\
                        &= 1 + P \dv{V}{U}
    \end{align}
    \todo{not sure where to go now}
}

@@

\question{3b. Particle flux expression}
\answer{
    Integration time! Basically just applying the double angle rule and doing some integrals.
    \begin{align}
        \phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta} \\
                       &= \frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \frac{1}{2} \int_0^{\pi/2} \sin (2 \theta) \dd{\theta} \\
                       &= -\frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} -\frac{1}{4} \qty[\cos (2 \theta)]_0^{\pi/2} \\
                       &= -\frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} -\frac{1}{4} \cdot \qty(-1 - 1) \\
                       &= \frac{v f(v)}{8 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \\
        \Aboxed{ \phi(v) \dd{v}  &= \frac{v f(v)}{4} \dd{v} }
    \end{align}
}

\question{3c. Total particle flux}
\answer{
    Right, let's see if I can remember how to do integrals! Let's start by performing a substitution.
    \begin{align}
        \phi &= \int_{\rm v_{\rm esc}}^{\infty} \phi(v) \dd{v} \\
             &= \frac{1}{4} \int_{\rm v_{\rm esc}}^{\infty} v f(v) \dd{v} \\
             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} \qty( \frac{m}{2 k_B T} )^{3/2} v^3 \exp \qty(- \frac{m v^2}{2 k_B T}) \dd{v} \\
             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \dd{v} \\
           u &= v^2, \qquad \dv{u}{v} = 2 v \\
        \phi &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} \frac{1}{2} A^{3/2} u \exp \qty(- A u) \dd{u} \\
             &= \sqrt{\frac{A^3}{32 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} u \exp \qty(- A u) \dd{u}
    \end{align}
    Okay let's pause for a second. That covers the substitution part, now we need to do integration by parts. Since I'm already using $u$ let's call the two variables in integration by parts $a$ and $b$ such that
    \begin{equation}
        \int a \dd{b} = a b - \int b \dd{a}
    \end{equation}
    In our case, we have that
    \begin{equation}
        a = u,\qquad \dd{b} = \exp(- A u)
    \end{equation}
    and thus we can also write that
    \begin{equation}
        \dd{a} = 1,\qquad b = - \frac{1}{A} \exp(- A u)
    \end{equation}
    So now putting that all together, we can write that the integral in full is
    \begin{align}
        &= \qty[ u \cdot - \frac{1}{A} \exp(- A u)]_{\rm v_{\rm esc}^2}^{\infty} - \int_{\rm v_{\rm esc}^2}^{\infty} - \frac{1}{A} \exp(- A u) \dd{u} \\
        &= \qty[ - \frac{u}{A} \exp(- A u)]_{\rm v_{\rm esc}^2}^{\infty} + \frac{1}{A} \int_{\rm v_{\rm esc}^2}^{\infty} \exp(- A u) \dd{u} \\
        &= \qty(0 + \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2)) - \qty[\frac{\exp(- A u)}{A^2}]_{\rm v_{\rm esc}^2}^{\infty} \\
        &= \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2}
    \end{align}
    Okay phew, we made it. Now let's go and plug that back in!
    \begin{align}
        \phi &= \sqrt{\frac{A^3}{32 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
             &= \frac{(1 + A v_{\rm esc}^2) e^{- A v_{\rm esc}^2}}{\sqrt{32 \pi A}}
    \end{align}
    Now let's put this into the format that you asked for. My constant $A$ is related to $v_s$ as $A = v_s^{-2}$ and thus we can write that $A v_{\rm esc}^2 = \lambda_{\rm esc}$. Additionally, we want this in terms of the total number of particles and so we multiply by the number density.
    \begin{equation}
        \boxed{ \phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{\sqrt{32 \pi}} }
    \end{equation}
}

>>>>

\question{\textbf{1. Heat Capacities}}
\answer{
    The definition of heat capacity gives that
    \begin{equation}
        C \equiv \dv{Q}{T}
    \end{equation}
    $C_P$ and $C_V$ are the same as $C$ but with the pressure, $P$, and volume, $V$, each held constant respectively. Applying the first law of thermodynamics, this gives
    \begin{equation}
        C_V = \qty(\dv{U}{T})_{V, N}\qquad C_P = \qty(\dv{U}{T})_{P} + P \qty(\dv{V}{T})_{P}
    \end{equation}
    Now let's apply the equipartition theorem and therefore write that
    \begin{equation}
        U = \frac{1}{2} N f k_B T
    \end{equation}
    where $f$ is the number of degrees of freedom. Now let's do some derivatives and apply the ideal gas law ($V = N k_B T / P$)
    \begin{align}
        C_{V} &= \dv{T} \qty(\frac{1}{2} N f k_B T)\\
              &= N k_B \cdot \frac{f}{2} \\
        C_{P} &= \dv{T} \qty(\frac{1}{2} N f k_B T) + P \dv{T} \qty(\frac{N k_B T}{P})\\
              &= \frac{1}{2} N f k_B + N k_B \\
              &= N k_B \qty(1 + \frac{f}{2}) \\
    \end{align}
    Therefore, the difference between the two is
    \begin{equation}
        \boxed{ C_P - C_V = N k_B }
    \end{equation}
    And dividing the two gives the adiabatic index
    \begin{equation}
        \boxed{ \frac{C_P}{C_V} = 1 + \frac{2}{f} = \gamma }
    \end{equation}
}

@@

\question{3b. Particle flux expression}
\answer{
    Integration time! Basically just applying the double angle rule and doing some integrals.
    \begin{align}
        \Phi(v) \dd{v} &= \frac{f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \int_0^{\pi/2} v \cos \theta \sin \theta \dd{\theta} \\
                       &= \frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \frac{1}{2} \int_0^{\pi/2} \sin (2 \theta) \dd{\theta} \\
                       &= -\frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} -\frac{1}{4} \qty[\cos (2 \theta)]_0^{\pi/2} \\
                       &= -\frac{v f(v)}{4 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} -\frac{1}{4} \cdot \qty(-1 - 1) \\
                       &= \frac{v f(v)}{8 \pi} \dd{v} \int_0^{2\pi} \dd{\phi} \\
        \Aboxed{ \Phi(v) \dd{v}  &= \frac{v f(v)}{4} \dd{v} }
    \end{align}
}

\question{3c. Total particle flux}
\answer{
    Right, let's see if I can remember how to do integrals! Let's start by performing a substitution.
    \begin{align}
        \Phi &= \int_{\rm v_{\rm esc}}^{\infty} \Phi(v) \dd{v} \\
             &= \frac{1}{4} \int_{\rm v_{\rm esc}}^{\infty} v f(v) \dd{v} \\
             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} \qty( \frac{m}{2 k_B T} )^{3/2} v^3 \exp \qty(- \frac{m v^2}{2 k_B T}) \dd{v} \\
             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \dd{v} \\
           u &= v^2, \qquad \dv{u}{v} = 2 v \\
        \Phi &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} A^{3/2} v^3 \exp \qty(- A v^2) \frac{\dd{u}}{2 v} \\
             &= \sqrt{\frac{1}{8 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} \frac{1}{2} A^{3/2} u \exp \qty(- A u) \dd{u} \\
             &= \sqrt{\frac{A^3}{32 \pi}} \int_{\rm v_{\rm esc}^2}^{\infty} u \exp \qty(- A u) \dd{u}
    \end{align}
    Okay let's pause for a second. That covers the substitution part, now we need to do integration by parts. Since I'm already using $u$ let's call the two variables in integration by parts $a$ and $b$ such that
    \begin{equation}
        \int a \dd{b} = a b - \int b \dd{a}
    \end{equation}
    In our case, we have that
    \begin{equation}
        a = u,\qquad \dd{b} = \exp(- A u)
    \end{equation}
    and thus we can also write that
    \begin{equation}
        \dd{a} = 1,\qquad b = - \frac{1}{A} \exp(- A u)
    \end{equation}
    So now putting that all together, we can write that the integral in full is
    \begin{align}
        &= \qty[ u \cdot - \frac{1}{A} \exp(- A u)]_{\rm v_{\rm esc}^2}^{\infty} - \int_{\rm v_{\rm esc}^2}^{\infty} - \frac{1}{A} \exp(- A u) \dd{u} \\
        &= \qty[ - \frac{u}{A} \exp(- A u)]_{\rm v_{\rm esc}^2}^{\infty} + \frac{1}{A} \int_{\rm v_{\rm esc}^2}^{\infty} \exp(- A u) \dd{u} \\
        &= \qty(0 + \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2)) - \qty[\frac{\exp(- A u)}{A^2}]_{\rm v_{\rm esc}^2}^{\infty} \\
        &= \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2}
    \end{align}
    Okay phew, we made it. Now let's go and plug that back in!
    \begin{align}
        \Phi &= \sqrt{\frac{A^3}{32 \pi}} \qty( \frac{v_{\rm esc}^2}{A} \exp(- A v_{\rm esc}^2) + \frac{\exp(- A v_{\rm esc}^2)}{A^2} ) \\
             &= \frac{(1 + A v_{\rm esc}^2) e^{- A v_{\rm esc}^2}}{\sqrt{32 \pi A}}
    \end{align}
    Now let's put this into the format that you asked for. My constant $A$ is related to $v_s$ as $A = v_s^{-2}$ and thus we can write that $A v_{\rm esc}^2 = \lambda_{\rm esc}$. Additionally, we want this in terms of the total number of particles and so we multiply by the number density.
    \begin{equation}
        \boxed{ \Phi = \frac{n_{\rm esc} v_s (1 + \lambda_{\rm esc}) e^{- \lambda_{\rm esc}}}{\sqrt{32 \pi}} }
    \end{equation}
}
